---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "qiwenb"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

---

## The question
For **Illinois**, I've chosen to examine **median home value** (variable B25077_001). Housing markets vary dramatically across the state, from Chicago's urban core to rural farming counties, but ACS estimates in sparsely populated areas often come with wider margins of error. I want to assess not only where property values are highest, but also which counties' estimates are statistically reliable enough to inform housing policy or investment decisions.

---

## Choosing what to measure


```{r}
my_state <- "IL"        # two-letter abbreviation, in quotes
my_variable <- "B25077_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```


**Illinois holds 102 counties.**



## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```


Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```


**Hardin County has the largest margin of error relative to its estimate (34.10%) in Illinois, which is the least reliable**

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 3  ~ "High confidence",
    moe_pct < 7 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```


**Based on the summary statistics, since the median is 5.27% and the 3rd quartile is 7.09%, the original 3%/6% thresholds would put most counties in "Moderate" or "Low". As a result, 21 counties are classified as "High confidence", 54 as "Moderate" and 27 as "Low confidence".**

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 3  ~ "High confidence",
    moe_pct < 7 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```



**High confidence counties have an average population 45 times larger than Low confidence counties (514,207 vs 11,301), confirming that smaller counties yield less reliable ACS estimates. While 53% of counties are moderately reliable, the 27 Low confidence counties—all sparsely populated—would be risky to use for targeted policy decisions without additional data collection.**

---

## What this means for policy


**What: The communities we know the least about in Illinois are its smallest, most rural counties. From my analysis, the 27 Low confidence counties—which include sparsely populated areas like Hardin, Pope, Calhoun, and Gallatin counties—have an average population of just 11,301 residents. These are precisely the places where ACS estimates are most unreliable, yet they represent a significant portion of Illinois's geographic area.**

**Why: The survey is sample‑based, not a census, and reliability depends heavily on sample size. Smaller counties contribute fewer respondents, which produces wider margins of error around every estimate. The ACS's own design prioritizes precision for larger populations, leaving rural communities with statistically noisy data.**

**Who: If a state agency distributing housing grants based on median home values, treating every county's estimate as equally certain. Using the 27 Low confidence counties' numbers at face value could misdirect funds—a county with an artificially high estimate might receive less aid than it actually needs, while another with a low estimate might be overlooked despite hidden poverty. The rural poor, already underserved, would be the ones hurt most.**

---

## Optional challenge: one function, five states

The county-level pull earlier in this report is hard-coded to one state. I wrapped
it in a function, `pull_county_reliability()`, that takes a state abbreviation (plus
optional variable and year) and returns each county's estimate, margin of error,
population, and `moe_pct`:

```{r}
pull_county_reliability <- function(state_abbr, variable = "B25077_001", year = 2023) {
  get_acs(
    geography = "county",
    variables = c(myvar = variable, pop = "B01003_001"),
    state = state_abbr,
    year = year,
    survey = "acs5",
    output = "wide"
  ) %>%
    transmute(
      state = state_abbr,
      GEOID,
      NAME,
      estimate = myvarE,
      moe = myvarM,
      popE,
      moe_pct = myvarM / myvarE * 100
    )
}
```

With `purrr::map_dfr()` the same analysis runs across several states at once. I
picked five with very different settlement geographies: Illinois, Ohio, Georgia,
Montana, and Wyoming — 428 counties in total.

```{r}
states <- c("IL", "OH", "GA", "MT", "WY")

multi_state <- map_dfr(states, pull_county_reliability)

multi_state %>%
  mutate(reliability = case_when(
    moe_pct < 3  ~ "High confidence",
    moe_pct < 7  ~ "Moderate",
    TRUE         ~ "Low confidence"
  )) %>%
  group_by(state, reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE),
    .groups = "drop"
  )
```

To measure the small-places-are-noisier pattern directly, I correlated county
population (log scale) with `moe_pct` within each state:

```{r}
multi_state %>%
  group_by(state) %>%
  summarize(correlation = cor(log(popE), moe_pct))
```

```{r}
ggplot(multi_state, aes(popE, moe_pct, color = state)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 7, linetype = "dashed", color = "grey50") +
  scale_x_log10(labels = scales::label_comma()) +
  labs(
    title = "Smaller counties, wider margins of error — in every state",
    x = "County population (log scale)",
    y = "Margin of error as % of estimate",
    color = "State"
  )
```

**What I found:** the small-places-are-noisier pattern holds everywhere. In all
five states the correlation between population and `moe_pct` falls in a narrow band
between -0.64 and -0.68, and the average High-confidence county is anywhere from
8× (Wyoming) to 45× (Illinois) more populous than the average Low-confidence county.

What differs between states is how much of the state is affected. In Montana, 40 of
56 counties and in Georgia, 93 of 159 fall into Low confidence, while in Ohio only
10 of 88 do. The *rule* is universal, but its footprint depends on settlement
geography: states dominated by small rural counties simply have less trustworthy
home-value data overall than states of evenly mid-sized counties like Ohio.
